Synthesizing Quotient Lenses

### 2017-10-18 papers/QuotientLensSynth/main.tex
More rewriting. Started on related work

Lenses are desirable since they save the programmer from having to write two
seperate transformations rather than one. Also, lenses offer strong
guarantees about how the two transformations interact since lenses are
constrained by lens laws.

However, these constraints are sometimes too strict. For example,  the
programmer may want the \bibtex{} to EndNote transformation to allow for
whitespace to occur arbitrarily between the words in the label, author and
title fields as well as in between them, and ignore the order in which the
author and title fields occur in each record. These rules allow for the
\bibtex{} citations
\begin{verbatim}
    @Book {conway,
      Author = "J. H. Conway",
      Title = {Regular Algebra and Finite Machines},
    }
\end{verbatim}
and
\begin{verbatim}
    @Book{     conway,
        Title = {Regular Algebra 
                 and Finite Machines},
           Author = "J. H. Conway",     }
\end{verbatim}
to be considered the same, with the EndNote records having similar rules
defined on them.

As a solution to this problem, Foster et al introduced {\em quotient lenses},
which are a generalization of lenses in which the source and target data are
considered modulo an equivalence relation. Quotient lenses are
both simpler and more flexible, since whole equivalence classes of data
values can be handled all at once and inessential details like whitespace
and ordering of fields can be ignored.

However many complicated equivalence relations are difficult to express in
Boomerang, the lens programming language introduced by Foster et al which
has quotient lens capability. For instance, while it is possible to express the
equivalence on the \bibtex{} data above in Boomerang, actually doing so is
difficult. The difficulty arises from the fact that the entire equivalence is
composed of three separate equivalences which nest and interact with each other
in a way that makes them difficult to express with the standard tools
available. Moreover, many of the equivalences expressible in Boomerang require
that the programmer first specify the data itself in form of a regular
expression before specifying the equivalence relation on the data.

@@

\begin{figure}[t]
\begin{lstlisting}
let wsp_to_space = project [ \n\t\r]$^+$ $\to$ " "
let valid_name = name | (name . wsp_to_space . (name . wsp_to_space)$^*$ . name)
let wsp_to_and = project ([ \n\t\r]$^+$ . "and" . [ \n\t\r]$^+$) $\to$ " and "
let bib_names = valid_name | (valid_name . wsp_to_and . (valid_name . wsp_to_and)$^*$ . valid_name)
let bib_author = "author = \"" . bib_names . "\""
let title = word | (word . wsp_to_space . (word . wsp_to_space)$^*$ . word)
let bib_title = "title = {" . title . "}"
let bib_perm = perm (label, bib_author, bib_title) with(project ("," . [ \n\t\r]$^*$ ) $\to$ ",\n")
let bibtex = "@book{" . bib_perm . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records. }
  \label{fig:example-qre}
\end{figure}

@@

\begin{lstlisting}
let bib_to_end : (lens in bibtex <=> endnot) 
= synth bibtex <=> endnote  using  {(bib_example, end_example)}
\end{lstlisting}
\noindent to synthesize a quotient lens that maps between \cd{bibtex} and
\cd{endnote} QREs, and sends the equivalence class of \cd{bib_bexample} to
the equivalence class of \cd{end_example}.

@@

\bcp{Make sure to list/discuss:
  \begin{itemize}
  \item Seminar papers on lenses (to which we can refer readers for further
  reading about connections to view updates, etc.) and quotient lenses.
  \item Discuss what other lens-like programming languages (besides
  boomerang) do about quotienting.  E.g., Augeas, XSugar, ...
  \item How quotient lenses are used in Boomerang.
  \item Search for papers that cite the quotient lens paper and see if they
  extend the ideas there.
  \item Search for theoretical papers involving equivalence relations over
  regular languages. 
  \end{itemize}
}

>>>>

Lenses are desirable since they save the programmer from having to write two
seperate transformations rather than one. Also, lenses offer strong
guarantees about how the two transformations interact since lenses are
constrained by lens laws. However, these constraints are sometimes too strict.
For example, the data may be such that a record can be represented in many ways
because of the presence of spurious whitespace characters or reordering of
fields. In such cases, deriving a lens which obeys the lens laws becomes a
daunting task.

As a solution to this problem, Foster et al proposed {\em quotient lenses},
which are a generalization of lenses in which the source and target data are
considered modulo an equivalence relation. Quotient lenses are
both simpler and more flexible, since whole equivalence classes of data
values can be handled all at once and inessential details like whitespace
and ordering of fields can be ignored.

However many complicated equivalence relations are difficult to express in
Boomerang, the lens programming language in which Foster et al implemented
quotient lenses. For example, suppose that the \bibtex{} to EndNote
transformation is required to allow for whitespace to occur arbitrarily between
the words in the label, author and title fields as well as in between them, and
ignore the order in which the author and title fields occur in each record.
These rules allow for the \bibtex{} citations
\begin{verbatim}
    @Book {conway,
      Author = "J. H. Conway",
      Title = {Regular Algebra and Finite Machines},
    }
\end{verbatim}
and
\begin{verbatim}
    @Book{     conway,
        Title = {Regular Algebra 
                 and Finite Machines},
           Author = "J. H. Conway",     }
\end{verbatim}
to be considered the same, with the EndNote records having similar rules
defined on them.

While it is possible to express the equivalence on the \bibtex{} data above in
Boomerang, actually doing so is difficult. The difficulty arises from the fact
that the entire equivalence is composed of three separate equivalences which
nest and interact with each other in a way that makes them difficult to express
with the standard tools available. Moreover, many of the equivalences
expressible in Boomerang require that the programmer first specify the data
itself in form of a regular expression before specifying the equivalence
relation on the data.

@@

\begin{figure}[t]
\begin{lstlisting}
let wsp = [ \n\t\r]$^+$
let wsp_to_space = project wsp $\to$ " "
let valid_name = name | (name . wsp_to_space . (name . wsp_to_space)$^*$ . name)
let wsp_to_and = project (wsp . "and" . wsp) $\to$ " and "
let bib_names = valid_name | (valid_name . wsp_to_and . (valid_name . wsp_to_and)$^*$ . valid_name)
let bib_author = "author = \"" . bib_names . "\""
let title = word | (word . wsp_to_space . (word . wsp_to_space)$^*$ . word)
let bib_title = "title = {" . title . "}"
let bib_perm = perm (label, bib_author, bib_title) with (project ("," .wsp) $\to$ ",\n")
let bibtex = "@book{" . bib_perm . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records. }
  \label{fig:example-qre}
\end{figure}

@@

\begin{lstlisting}
let bib_to_end : (lens in bibtex <=> endnote) 
= synth bibtex <=> endnote  using  {(bib_example, end_example)}
\end{lstlisting}
\noindent to synthesize a quotient lens that maps between \cd{bibtex} and
\cd{endnote} QREs, and sends the equivalence class of \cd{bib_bexample} to
the equivalence class of \cd{end_example}.

@@

Bidirectional transformations have been used in diverse areas of computing
where they arise as parsers and pretty printers, marshallers
and unmarshallers, serializers and deserializers, database views and view
updaters and many others\sam{TODO put citations}. Such transformations have
been extensively studied since they were proposed as a solution to the classical
{\em view-update problem} in the database community, where the challenge is to
derive a program which extracts a view of data from a source, as well as a
program which folds view updates back into the source safely and correctly.

This paper builds on the work of Foster et al~\cite{quotientlenses} who
introduced the theory of quotient lenses and implemented quotient lenses as a
refinement of the bidirectional string processing language Boomerang.
In Boomerang, the source and target types are specified using regular
expressions, and the equivalence relations are expressed using canonizers, which
are functions that map elements of a regular language to their canonical
representative. Boomerang canonizers can express a very broad class of
equivalence relations between regular languages, but actually doing so is often
difficult with complicated equivalence relations. For instance, Boomerang's
in-build permutation combinator does not account for separators in between the
regular expressions. Also, this combinator permutes regular expressions rather
than canonizers, thus making it difficult to express nested permutations which
occur in many data formats, especially XML and XML-like formats.

In their paper, Foster et al also discuss other bidirectional programming
languages which support quotienting of data including XSugar~\cite{xsugar},
biXid~\cite{bixid} and X/Inv~\cite{xinv}. XSugar programs bidirectionally
convert data stored in XML and ASCII formats respectively, with the
transformations specified by a pair of unambiguous grammars. The quotienting
occurs on the XML side by use of a generic canonizer which standardizes the
representation of trees. Well-formed XSugar programs are guaranteed to be
bijective modulo an equivalence relation that captures XML normalization. biXid
programs convert between pairs of XML documents, with the XML formats specified
using a pair of grammars as in XSugar. However, biXid grammars can be ambiguous.
This ambiguity is what allows biXid to express equivalences on the data.
Finally, Foster et al discuss a possible connection with the language languages
X and Inv which support a primitive duplication combinator that does not work
well with the lens laws, but can be expressed using Boomerang quotient lenses.

 \bcp{Make sure to list/discuss:
  \begin{itemize}
  \item Seminar papers on lenses (to which we can refer readers for further
  reading about connections to view updates, etc.) and quotient lenses.
  \item Search for papers that cite the quotient lens paper and see if they
  extend the ideas there.
  \item Search for theoretical papers involving equivalence relations over
  regular languages. 
  \end{itemize}
}
